ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-09-27 rebuttal.tex

% bump up section font size
\usepackage{sectsty}
\sectionfont{\fontsize{12}{15}\selectfont}

\usepackage[colorlinks=true]{hyperref}
\usepackage{todonotes}

\setlength{\parindent}{0in}
\setlength{\parskip}{0.4cm}

\newenvironment{reviewer}{\vspace{0.5\baselineskip}\begingroup\itshape\textbf{Reviewer:}}{\endgroup\vspace{0.5\baselineskip}}
\newenvironment{response}{\vspace{0.5\baselineskip}\begin{response}

\end{response}}{\vspace{0.5\baselineskip}}
\newcommand{\review}[1]{{\itshape#1}}
\newcommand{\respond}[1]{\begin{response}

\end{response} #1}

@@

\begin{document}

We thank the reviewers for their helpful and detailed comments, and have addressed their remarks here.
The manuscript has been significantly revised, and we marked significant changes corresponding to suggestions from Reviewers \#1 and \#2 in \hlg{yellow} and \hly{green}, respectively.

\section*{Reviewer \#1:}

\begin{reviewer}
    %
    Currently four database formats are offered for storing combustion experimental data. These are
    CloudFlame (Excel/CSV format), PrIMe (XML-based format), ReSpecTh (also XML-based format), and
    ChemKED (YAML-based format). Elaboration of the last format is reported in the submitted
    manuscript. The motivation for the elaboration of this new format was based on the statement
    that YAML is more human readable than XML. I think that this is simply not true. Even a
    first-time user can read the raw datafiles of both formats with equal efficiency. Also,
    currently the suggested new format is much more limited (e.g. it is able to encode ignition
    delay time experiments only). The presently planned features of ChemKED are already available in
    the other three formats.

    However, I suggest the publication of an improved and shortened version of the manuscript in
    IJCK. The reason is that a permanent and laboratory-independent storage of experimental data is
    an important issue and I agree that a public discussion promotes the acceptance of a widely used
    common format.
    %
\end{reviewer}

\begin{response}
    %
    We would like to thank the reviewer for their comments. Detailed responses to each comment are
    below.
    %
\end{response}

\begin{reviewer}
    %
    I think that the paper contains too much information science details. Still, as the authors
    note, the description is not comprehensive and they refer to the user manual. I suggest the
    creation of a significantly abridged version that contains the basic information technology
    points only and does not discuss the details. For example, Chapter 4 could be deleted and
    Chapters 2 and 3 may also be shortened.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    p.~1 ``Competing standards such as the Cantera [2] CTI format or FlameMaster [3] lag behind
    considerably, although use of the former continually grows due to Cantera?s open availability.''

    Note that FlameMaster is also freely available from the GitLab repository of the developers.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    p.~2 ``Second, the XML format is intended to be a machine-readable markup language rather than a
    data format, and its lack of human readibility presents a barrier to creating and working with
    database files.''

    XML is a text format and it is fully human readable, similarly to YMAL. Even a
    first-time user is able to read the chemical content of a  PrIMe XML file and an experienced
    user is able to write such a file.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    I suggest adding to following block to this para:

    ``The original PrIMe was on an Amazon cloud server. Recently it was shifted to the computer
    center of DLR Stuttgart. At the new location it is called ``PrIMe 2.0''
    [\url{https://prime.cki-know.org/}]. This is identical to the original PrIMe apart from
    technical changes due to the different computation background. Professor Frenklach plans to
    develop further PrIMe (using name PrIMe 3.0) and this version will be availabe at
    [\url{http://primekinetics.org/}].''
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    ``For example, where PrIMe uses internal bibliographic references, ReSpecTh adds a field for
    typical bibliographic data.''

    I suggest the following extension:

    ``For example, where PrIMe uses internal bibliographic references, ReSpecTh adds a field for
    typical bibliographic data, including the DOI of the cited arecicle. Each datafile has a unique
    own DOI and unique URL address.
    [\url{http://respecth.chem.elte.hu/respecth/reac/CombustionData.php}]''. Also species can be
    identified by InCHI or SMILES.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    ``The ReSpecTh standard also provides machine-readable formats for describing ignition
    experiments, including a field for the definition of autoignition and the ability to specify
    facility-specific effects.''

    Modified text:

    ``The ReSpecTh standard provides formats that are both human and machine-readable for describing
    ignition experiments, including a field for the definition of autoignition and the ability to
    specify facility-specific effects. Also, it describes PSR and flow reactor speciation data,
    burner stabilized flame speciation data and burning velocity data.''
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    ``However, ReSpecTh experimental files are another XML-based format, and, as such, suffer from
    the same usability issues as PrIMe.''

    I think that the limits of an XML database is set by the actual set of tags and not the format.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    ``Moreover, the standard remains closed, requiring potential users to register with the
    standard's authors to access the specification and the growing database.''

    I suggest instead:

    ``Registration is open for everyone who provides a valid institutional e-mail address.''
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    ``... hydrogen/syngas [11] and ethanol [12]. In all four cases, they converted numerous
    experimental datasets from the literature into the ReSpecTh standard.''

    Extended version:

    ``... hydrogen/syngas [11], methanol and formaldehyde [C. Olm, T. Varga, E. Valko, H. J. Curran,
    T. Turanyi: Uncertainty quantification of a newly optimized methanol and formaldehyde combustion
    mechanism, Combustion and Flame, in press] and ethanol [12]. In all five cases, they converted
    numerous experimental datasets from the literature into the ReSpecTh standard. The database
    currently contains  93240 experimental data points in 1376 XML-format data files
    [\url{http://respecth.chem.elte.hu/respecth/reac/CombustionData.php}].''
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    ``type (required, string): How ignition delay was measured;''

    The options described here is a very limited subset of the ignition delay time definitions used
    in experimental articles. For example, backward or forward interpolated values of a measured
    quantity are further possibilities.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    p.~8 ``In the latter case, some information may be lost, as ChemKED files both encode more
    details and specify more granularity of some information than ReSpecTh files.''

    This was true for ReSpecTh format 1.0, but no longer true for ReSpecTh Kinetic Dataformat 2.0.
    The latter covers all info of the ChemKED files (and more).
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    Missing data in the references: Year is missing is refs. [5], [6], [9], [10], [11], [12],
    [21],[25],[27].
    %
\end{reviewer}

\begin{reviewer}
    %
    [3] Please add reference: \url{https://www.itv.rwth-aachen.de/index.php?id=13}  Also, the
    latest version is from 2017.

    [7] Further info:  Paper P1-04, ISBN 978-963-12-1257-0

    [8] Please refer to ReSpecTh Kinetics Data Format Specification 2.0 instead.
    %
\end{reviewer}

\begin{response}

\end{response}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Reviewer \#2:}

\begin{reviewer}
    %
    My main concern with the creation of a publically available database, formatted with ChemKED, is
    that it seems that the creation of the files has to be done manually, i.e. susceptible for
    errors. This is especially true if the experiment comprises a big set of datapoints and/or
    hundreds of measured species. Furthermore, the YAML data serialization format is indeed
    human-readable, but it does not provide an easy summary/overview of the experimental data, such
    as an excel or CSV file would give, and therefore, people might be not willing to put the extra
    effort.
    %
\end{reviewer}

\begin{reviewer}
    %
    1.~How are the authors planning on `forcing' the experimentalists to publish their data in this
    new ChemKED file format?
    %
\end{reviewer}

\begin{response}
    %
    While we cannot force anyone to use a particular format, we hope that by defining a format and
    ecosystem with desirable features---including inherent openness and usability---experimentalists
    will begin adopting a community-standard format. In fact, we have already received interest from
    various groups, and hope this will grow as we continue to improve and add features. We also hope
    that at a certain point, with wide adoption, sharing experimental data in a standard format will
    become the expected norm (e.g., when submitting an article for publication) like sharing a
    Chemkin-format chemical kinetic model.
    %
\end{response}

\begin{reviewer}
    %
    2.~Am I correct to state that ChemKED is a YAML file with a predefined structure?
    %
\end{reviewer}

\begin{response}
    %
    Yes, ChemKED files are files in YAML format, with a prescribed structure and set of contents.
    %
\end{response}

\begin{reviewer}
    %
    3.~I think the article would be more appealing if it would have at least one figure as an
    example.
    %
\end{reviewer}

\begin{response}

\end{response}

\subsection*{Abstract and introduction}

\begin{reviewer}
    %
    4. In the abstract, it is mentioned that the Python-based package PyKED creates ChemKED-
    formatted files. However in the article there is no evidence that PyKED aides in creating
    ChemKED-formatted files.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    5.~Page 1, line 31: The future development plans include support for jet stirred reactors,
    however, this is not mentioned in section 5, Conclusions and future work, nor is there any
    reference to it on the GitHub wiki roadmap of PyKED.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
        6. I don't completely agree with the second part of the statement on page 2 line 16-18:
        ``Second, the XML format is intended to be a machine-readable mark-up language rather than a
        data format, and its lack of human readability presents a barrier to creating and working
        with database files.''

        In my opinion the YAML format resembles the XML format strongly. XML format is human-readable,
        however, because of the html-tags and `\texttt{<>} characters, the YAML format might be easier
        to read. If XML would lack human readability as it is stated in the article, it would not be
        such a popular format, and I don't think PrIMe and ReSpecTh would be using it.
    %
\end{reviewer}

\begin{response}
    %
    We agree with the reviewer that XML is technically human readable by an expert (as opposed to,
    say, a binary format), and also that YAML is easier to read---by ``human-readable'', we really
    mean human-\emph{understandable}.
    %
\end{response}

\begin{reviewer}
    %
    7.~An initial version of the ChemKED format was introduced by Niemeyer, as is mentioned in the
    introduction. The mentioned article describes a Python-package called PyTeCK. How does this
    PyTeCK package relate to the PyKED package that is presented in the submitted article?
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
%
8.~Page 3, line 10: ``Our motivations are similar to those of '', `to' is missing.
%
\end{reviewer}

\begin{response}
    %
    We have edited that sentence to correct the error.
    %
\end{response}

@@

\subsection*{Overview of ChemKED format}

\begin{reviewer}
    %
    9.~A reference should not be a required mapping of an experimental dataset. If the intention is
    to make the community to adopt the ChemKED file structure, it might be easier to do so if
    experimentalists can use the structure internally as well, i.e. to read, store, and share
    non-published internal data.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    10.~Page 4, line 32: ``The second section of the file encodes the experimental data''. Perhaps
    it is easier to follow if this is briefly mentioned before the authors explain the details of
    the first section of the ChemKED format.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    11.~Why do both \texttt{atomic-composition} and \texttt{elemental-composition} exist as
    synonyms? Wouldn't it be clearer if only one of the two can be used as a valid input?
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    12.~Is it possible to create for example two \texttt{common-properties} \texttt{composition}
    blocks A and B and refer to them in the \texttt{DataPoint} keys?
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    13.~Page 6, line 12 describes the initial efforts of building an open repository of ChemKED
    files, on GitHub. Is this the most efficient way for making a publically available database? One
    issue with this is the fact that adding files is done via pull requests, which have to be
    verified and approved by the owners of the repository. Therefore a certain delay is inevitable.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    14.~Page 6, line 18 states that a converter exists from ReSpecTh to ChemKED. But isn't this
    feature added when going from v0.1.6 to v0.2.0, which makes page 3 line 21: v0.1.6 not correct.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    15.~Page 6, line 21, is the `,' necessary after fork?
    %
\end{reviewer}

\begin{response}
    %
    We removed the extraneous comma.
    %
\end{response}

@@

\subsection*{PyKED architecture}

\begin{reviewer}
    %
    16.~The Python package Cerberus is used to validate the format and content of the given ChemKED
    file. How does this package handles/reports the inconsistencies with the schemas? As far as my
    local test went with giving a corrupted YAML file to the ChemKED initiator, the package just
    spits out a Python error, which might be not the most user-friendly solution.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    17.~A recurring issue with actively developing codes is backwards compatibility. How are the
    authors planning on addressing this issue?
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    18.~Page 7, line 12, \texttt{mole-percent}, \texttt{mole-fraction}, and \texttt{mass-fraction}
    include a dash in their name, whereas this is not the case for the example on page 5 line 48 and
    in the description of the composition element on page 4, line 47.

    19.~Same issue as \#18 holds for page 7, line 17.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    20.~I think it would be more clear if \texttt{volume\_history}, \texttt{compression\_time},
    \texttt{compressed\_temperature}, and \texttt{compressed\_pressure} would be combined.

    The line: `if the ChemKED file encodes an RCM experiment' is unnecessarily repeated.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    21.~Page 7, line 39: ``Future versions of PyKED may support this feature via, e.g. online
    lookup''. There are two main caveats with online lookups: a) It will make the code slower if the
    internet connection is not strong, b) an internet connection is necessary for the code to be
    executed.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    22.~Page 7, line 40: Why is it interesting to create the YAML file again with the
    \texttt{write\_file()} function. Creating a summary excel/CSV file would be more useful.
    %
\end{reviewer}

\begin{response}

\end{response}

@@

\subsection*{Usage examples}
\subsubsection*{RCM modelling with varying reactor volume}

\begin{reviewer}
    %
    23.~Page 10, line 29--30: ``A user might want to load the information'' Avoid the term `might'.
    It's one of the main goals of using ChemKED and its associated Python-package PyKED to load the
    information necessary for simulating the experiment.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    24.~The Python code snippets are not always initiated in the text.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    25.~Page 10, line 55: add more commentary text. It is not clear that those lines create the
    necessary instances for Cantera, rather than ChemKED.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    26.~Page 11, line 7: Where is this `air.xml' file coming from. This is not clear from the text.
    %
\end{reviewer}

\begin{response}

\end{response}

@@

\subsubsection*{Shock tube modelling with constant volume}

\begin{reviewer}
    %
    27.~The table overlapping page 11 and 12 can be reduced by truncating the \texttt{DataPoints},
    similarly as is done with the table starting at page 9, line 31 and ending at page 10, line 27.
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    28. Python code snippet on page 13: (a.) Add more commentary text. (b.) Changing the example to
    enable multithreading might make it more appealing to start using ChemKED as a data format in
    addition with Cantera as a simulation tool.
    %
\end{reviewer}

\begin{response}

\end{response}

@@

\subsection*{Conclusions and future work}

\begin{reviewer}
    %
    29.~No conclusion is mentioned for the usage examples.
    %
\end{reviewer}

\begin{response}

\end{response}

### 2017-09-27 rebuttal.tex

% bump up section font size
\usepackage{sectsty}
\sectionfont{\fontsize{12}{15}\selectfont}

\usepackage[colorlinks=true]{hyperref}
\usepackage{todonotes}

\setlength{\parindent}{0in}
\setlength{\parskip}{0.4cm}

\newenvironment{reviewer}{\vspace{0.5\baselineskip}\begingroup\itshape\textbf{Reviewer:}}{\endgroup\vspace{0.5\baselineskip}}
\newenvironment{response}{\vspace{0.5\baselineskip}\textbf{Our Response:}}{\vspace{0.5\baselineskip}}
\newcommand{\review}[1]{{\itshape#1}}
\newcommand{\respond}[1]{\textbf{Our Response:} #1}

@@

    This was true for ReSpecTh format 1.0, but no longer true for ReSpecTh Kinetic Dataformat 2.0.
    The latter covers all info of the ChemKED files (and more).
    %
\end{reviewer}

\begin{response}

\end{response}

\begin{reviewer}
    %
    Missing data in the references: Year is missing is refs.~[5], [6], [9], [10], [11], [12],
    [21],[25],[27].

    [3] Please add reference: \url{https://www.itv.rwth-aachen.de/index.php?id=13}  Also, the
    latest version is from 2017.

@@

    [7] Further info:  Paper P1-04, ISBN 978-963-12-1257-0

    [8] Please refer to ReSpecTh Kinetics Data Format Specification 2.0 instead.
    %
\end{reviewer}

\begin{response}
    We corrected those references; thank you for pointing out the missing\slash old information.
\end{response}

### 2017-09-27 pyked.tex
Additional ReSpecTh references to address #13

Nonetheless, databases of fundamental combustion experiments
have proved quite useful. For instance, Olm and colleagues used their
ReSpecTh-based database to quantify the performance of literature
hydrogen~\autocite{Olm:2014gn} and syngas~\autocite{Olm:2015ch} kinetic models,
and more recently develop improved models for hydrogen\slash
syngas~\autocite{Varga:2016gj}, methanol\slash formaldehyde~\autocite{Olm:2017a},
and ethanol~\autocite{Olm:2016et}. In all four
cases, they converted numerous experimental datasets from the literature into
the ReSpecTh standard.

>>>>

Nonetheless, databases of fundamental combustion experiments
have proved quite useful. For instance, Olm and colleagues used their
ReSpecTh-based database to quantify the performance of literature
hydrogen~\autocite{Olm:2014gn} and syngas~\autocite{Olm:2015ch} kinetic models,
and more recently develop improved models for hydrogen\slash
syngas~\autocite{Varga:2016gj}, methanol\slash formaldehyde~\autocite{Olm:2017a},
and ethanol~\autocite{Olm:2016et}. In all five cases, they converted numerous
experimental datasets from the literature into the ReSpecTh standard.
Their database currently contains \num{93240} experimental data points in 1376
ReSpecTh XML files~\autocite{respecth}.
